### 2020-10-15 conclusion.tex
Processed more feedback.

The groundbreaking work on progress for session-typed $\pi$-calculus, by Dezani \etal~\cite{dezani-ciancaglinimostrous06}, guarantees progress by allowing only one active session at a time. In~\cite{dezani-ciancagliniliguoro09progress} the authors introduce a partial order on channels in line with Kobayashi's work~\cite{kobayashi98}. Carbone and Debois~\cite{carbonedebois10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} used to provide a missing counterpart to a process, thus guaranteeing progress.
Carbone \etal~\cite{carbonedardha14} studied further the use of catalysers and showed that progress is a compositional form of lock-freedom. As in~\cite{padovani14} the authors show that by using the encoding of session types~\cite{dardhagiachino12} and Kobayashi's obligations/capabilities, we can obtain progress for session types. Vieira and Vasconcelos~\cite{vieiravasconcelos13} used single priorities and an abstract partial order to guarantee deadlock freedom in a session-typed $\pi$-calculus.

@@

\subsubsection*{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming~\cite{wadler15}. With the rise of linear logic~\cite{girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus~\cite{abramsky94,bellinscott94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning~\cite{cairespfenning10}, for dual intuitionistic linear logic (DILL), and later by Wadler~\cite{wadler15} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via the logical cut rule. Dardha and Gay follow Padovani's work~\cite{padovani14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic correspondence with session types~\cite{wadler12}. Dardha and P\'{e}rez~\cite{dardhaperez15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \etal study sharing~\cite{balzerpfenning17} and later on deadlock freedom~\cite{balzertoninho19} in a session-typed $\pi$-calculus based on the correspondence with DILL. In~\cite{balzerpfenning17} types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. In~\cite{balzertoninho19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

>>>>

The groundbreaking work on progress for session-typed $\pi$-calculus, by Dezani \etal~\cite{dezani-ciancaglinimostrous06}, guarantees progress by allowing only one active session at a time. In~\cite{dezani-ciancagliniliguoro09progress} the authors introduce a partial order on channels in line with Kobayashi's work~\cite{kobayashi98}. Carbone and Debois~\cite{carbonedebois10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} used to provide a missing counterpart to a process, thus guaranteeing progress.
Carbone \etal~\cite{carbonedardha14} studied further the use of catalysers and showed that progress is a compositional form of lock-freedom. Following Padovani~\cite{padovani14}, the authors show that by using the encoding of session types~\cite{dardhagiachino12} and Kobayashi's obligations/capabilities, we can obtain progress for session types. Vieira and Vasconcelos~\cite{vieiravasconcelos13} used single priorities and an abstract partial order to guarantee deadlock freedom in a session-typed $\pi$-calculus.

@@

\subsubsection*{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming~\cite{wadler15}. With the rise of linear logic~\cite{girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus~\cite{abramsky94,bellinscott94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning~\cite{cairespfenning10}, for dual intuitionistic linear logic (DILL), and later by Wadler~\cite{wadler15} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via the logical cut rule. Dardha and Gay follow Padovani's work~\cite{padovani14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic correspondence with session types~\cite{wadler12}. Dardha and P\'{e}rez~\cite{dardhaperez15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \etal study sharing~\cite{balzerpfenning17} and later on deadlock freedom~\cite{balzertoninho19} in $\text{SILL}_S$ and $\text{SILL}_{S+}$, session-typed $\pi$-calculi based on the correspondence with DILL. In $\text{SILL}_S$~\cite{balzerpfenning17}, types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. In $\text{SILL}_{S+}$~\cite{balzertoninho19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

### 2020-10-15 introduction.tex
Processed more feedback.

With regards to GV, the original work on session types in GV~\cite{gayvasconcelos10,gayvasconcelos12} did not satisfy deadlock freedom. Later on, calculi in the GV family~\cite{wadler15,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, \ie by combining channel creation and thread spawning into a single operation, called fork. The fork construct is the term representation of the cut construct in CP, which evaluates to a name restriction combined with a parallel composition. Unfortunately, as with CP, this is overly restrictive, as it limits the possible communication structures to trees. There are several benefits of working with a functional language as opposed to process calculi: a functional language supports higher-order functions (and therefore abstraction) not usually present in process calculi; it allows us to derive extensions of the communication fragment of the language via well-understood extensions of the functional fragment, \ie deriving internal/external choice from sum types; and in concurrent functional calculi, there is a strict separation between the program that the user writes, and the state of a system as it evaluation. In the setting of process calculi, these are conflated.

Moreover, the benefit of working with GV over other session-typed functional languages is that GV has strong ties to linear logic, via its relation to CP~\cite{wadler12}, and consequently it has strong formal properties, \eg deadlock freedom, albeit in a restrictive way.
Following the above observations, we aim to further investigate GV and deadlock freedom and thus pose our research question:

@@

We make the following contributions:
\begin{itemize}
\item \emph{Priority GV} We present Priority GV (PGV) in~\Cref{sec:pgv}; we present the syntax of types and terms, operational semantics and typing rules. We show our type system for PGV is sound by proving subject reduction (\Cref{thm:pgv-subject-reduction-confs}) and closed progress (\Cref{lem:pgv-closed-progress-confs}).
\item \emph{Milner's Cyclic Scheduler in PGV} We present, for the first time, Milner's Cyclic Scheduler~\cite{milner89} (\Cref{sec:pgv-example}) as a GV term. This example illustrates how a cyclic structure can now be expressed in PGV thanks to the separation of channel creation and thread creation, and priorities.
\item \emph{Updated Priority CP} We present an updated version of Priority CP (PCP)~\cite{dardhagay18} in \Cref{sec:pcp} where we remove the commuting conversions and move away from a reduction relation as cycle elimination, and towards a reduction as one would expect for a process calculus. We then show Milner's Cyclic Scheduler in PCP (in \Cref{sec:pcp-example}) originally presented in~\cite{dardhagay18}.
\item \emph{Encoding PCP into PGV} We present an encoding of PCP into PGV in \Cref{sec:pcp-to-pgv}. We prove the encoding is sound and complete with respect to typing (\Cref{thm:pcp-to-pgv-confs-preservation}) and operational semantics (\Cref{thm:pcp-to-pgv-operational-correspondence-soundness} and \Cref{thm:pcp-to-pgv-operational-correspondence-completeness}). Finally, we can encode Milner's Cyclic Scheduler in PCP (\Cref{sec:pcp-example}) resulting in the Scheduler in PGV (\Cref{sec:pgv-example}).
\end{itemize}
\end{document}

>>>>

With regards to GV, the original work on session types in GV~\cite{gayvasconcelos10,gayvasconcelos12} did not satisfy deadlock freedom. Later on, calculi in the GV family~\cite{wadler15,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, \ie by combining channel creation and thread spawning into a single operation, called fork. The fork construct is the term representation of the cut construct in CP, which evaluates to a name restriction combined with a parallel composition. Unfortunately, as with CP, this is overly restrictive, as it limits the possible communication structures to trees. There are several benefits of working with a functional language as opposed to process calculi. Functional languages support higher-order functions, and have a capability for abstraction not usually present in process calculi. Working within a functional language allows us to derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \ie deriving internal/external choice via sum types. Finally, in concurrent functional calculi, there is a strict separation between the program that the user writes, and the state of a system as it evaluation. In the setting of process calculi, these are conflated.

Moreover, the benefit of working with GV over other session-typed functional languages is that GV has strong ties to linear logic, via its relation to CP~\cite{wadler12}, and consequently it has strong formal properties, \eg deadlock freedom in a restrictive way.
Following the above observations, we aim to further investigate GV and deadlock freedom and thus pose our research question:

@@

We make the following contributions:
\begin{itemize}
\item \emph{Priority GV} We present Priority GV (PGV) in~\Cref{sec:pgv}; we present the syntax of types and terms, operational semantics and typing rules. We show our type system for PGV is sound by proving subject reduction (\Cref{thm:pgv-subject-reduction-confs}) and closed progress (\Cref{lem:pgv-closed-progress-confs}).
\item \emph{Milner's Cyclic Scheduler in PGV} We present, for the first time, Milner's Cyclic Scheduler~\cite{milner89} (\Cref{sec:pgv-example}) as a GV term. This example illustrates how a cyclic structure can now be expressed in PGV thanks to the separation of channel creation and thread creation, and priorities.
\item \emph{Updated Priority CP} We present an updated version of Priority CP (PCP)~\cite{dardhagay18} in \Cref{sec:pcp} where we remove the commuting conversions and move away from a reduction relation as cycle elimination, and towards a reduction as one would expect for a process calculus. We then show Milner's Cyclic Scheduler in PCP (in \Cref{sec:pcp-example}) originally presented in~\cite{dardhagay18}.
\item \emph{Encoding PCP into PGV} We present an encoding of PCP into PGV in \Cref{sec:pcp}. We prove the encoding is sound and complete with respect to typing (\Cref{thm:pcp-to-pgv-confs-preservation}) and operational semantics (\Cref{thm:pcp-to-pgv-operational-correspondence-soundness} and \Cref{thm:pcp-to-pgv-operational-correspondence-completeness}). Finally, we can encode Milner's Cyclic Scheduler in PCP (\Cref{sec:pcp-example}) resulting in the Scheduler in PGV (\Cref{sec:pgv-example}).
\end{itemize}
\end{document}

### 2020-10-15 priority-gv.tex
Processed more feedback.

\subsubsection*{Subject reduction}
Unlike with previous versions of GV, structural congruence, term reduction, and configuration reduction are all type preserving, though it is not entirely trivial to show.

Specifically, we need to show that substitution preserves priority constraints. For this, it helps to prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication actions, and that any priorities present in the type of the value come from the typing environment.
\begin{restatablelemma}{lempgvvaluedone}
  \label{lem:pgv-value-done}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
\end{restatablelemma}
\begin{proof}
  By induction on the derivation of $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$. We demonstrate the case for \LabTirName{T-Inl}, as it illustrates the need for the side conditions on \LabTirName{T-Inl} and \LabTirName{T-Inr}:
  \begin{case*}[\LabTirName{T-Inl}]
    The induction hypothesis gives us $\cs{p}=\cs{\pbot}$, and $\pr(\ty{\Gamma})=\pr{(\ty{T})}$. We know $\pr(\ty{T})=\pr({\ty{U}})$, hence $\pr(\ty{\Gamma})=\pr(\ty{\tysum{T}{U}})$.
    \begin{mathpar}
      \small
      \inferrule*{
        \tseq[\cs{p}]{\ty{\Gamma}}{V}{T}
        \\
        \pr(\ty{T})=\pr(\ty{U})
      }{\tseq[\cs{p}]{\ty{\Gamma}}{\inl{V}}{\tysum{T}{U}}}
    \end{mathpar}
  \end{case*}
  The full proof can be found in~\cref{prf:lem-pgv-value-done}.
\end{proof}
\begin{restatablelemma}{lempgvsubstitution}[Substitution]
  \label{lem:pgv-substitution}
  \hfill\\%newline before theorem statement
  If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
\end{restatablelemma}
\begin{proof}
  By induction on the derivation of $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$. Whenever the contexts in the premises and conclusion differ, we use \cref{lem:pgv-value-done} to prove that the priorities are preserved. We demonstrate the case for \LabTirName{T-Lam}:
  \begin{case*}[\LabTirName{T-Lam}]
    By \cref{lem:pgv-value-done}, $\pr(\ty{\Theta})=\pr(\ty{U'})$, hence $\pr(\ty{\Gamma},\ty{\Theta})=\pr(\ty{\Gamma},\ty{U'})$.
    \begin{mathpar}
      \small
      \inferrule*{
        \tseq[\cs{q}]{\ty{\Gamma},\tmty{x}{U'},\tmty{y}{T}}{M}{U}
      }{\tseq[\cs{\pbot}]{\ty{\Gamma},\tmty{x}{U'}}
        {\lambda y.M}
        {\tylolli[\cs{\pr(\ty{\Gamma},\ty{U'})},\cs{q}]{T}{U}}}
      \substarrow{V}{x}
      \inferrule*{
        \tseq[\cs{q}]{\ty{\Gamma},\ty{\Theta},\tmty{y}{T}}{\subst{M}{V}{x}}{U}
      }{\tseq[\cs{\pbot}]{\ty{\Gamma},\ty{\Theta}}
        {\lambda y.\subst{M}{V}{x}}
        {\tylolli[\cs{\pr(\ty{\Gamma},\ty{\Theta})},\cs{q}]{T}{U}}}
    \end{mathpar}
  \end{case*}
  The complete proof can be found in~\cref{prf:lem-pgv-substitution}.
\end{proof}
The proofs for subject congruence and subject reduction are standard inductions. The details can be found in \cref{prf:lem-pgv-subject-reduction-terms,prf:lem-pgv-subject-congruence,prf:thm-pgv-subject-reduction-confs}.
\begin{restatablelemma}{lempgvsubjectreductionterms}[Subject Reduction, $\tred$]
  \label{lem:pgv-subject-reduction-terms}
  \hfill\\%newline before theorem statement
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
  then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
\end{restatablelemma}
\begin{restatablelemma}{lempgvsubjectcongruence}[Subject Congruence, $\equiv$]
  \label{lem:pgv-subject-congruence}
  \hfill\\%newline before theorem statement
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{restatablelemma}
\begin{restatabletheorem}{thmpgvsubjectreductionconfs}[Subject Reduction, $\cred$]
  \label{thm:pgv-subject-reduction-confs}
  \hfill\\%newline before theorem statement
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{restatabletheorem}

>>>>

\subsubsection*{Subject reduction}
Unlike with previous versions of GV, structural congruence, term reduction, and configuration reduction are all type preserving.

Most of the subject reduction proof is standard. Novel is that we must show that substitution preserves priority constraints. For this, it helps to prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication actions, and that any priorities present in the type of the value come from the typing environment.
\begin{restatablelemma}{lempgvvaluedone}
  \label{lem:pgv-value-done}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
\end{restatablelemma}
\begin{proof}
  By induction on the derivation of $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$. We demonstrate the case for \LabTirName{T-Inl}, as it illustrates the need for the side conditions on \LabTirName{T-Inl} and \LabTirName{T-Inr}:
  \begin{case*}[\LabTirName{T-Inl}]
    The induction hypothesis gives us $\cs{p}=\cs{\pbot}$, and $\pr(\ty{\Gamma})=\pr{(\ty{T})}$. We know $\pr(\ty{T})=\pr({\ty{U}})$, hence $\pr(\ty{\Gamma})=\pr(\ty{\tysum{T}{U}})$.
    \begin{mathpar}
      \small
      \inferrule*{
        \tseq[\cs{p}]{\ty{\Gamma}}{V}{T}
        \\
        \pr(\ty{T})=\pr(\ty{U})
      }{\tseq[\cs{p}]{\ty{\Gamma}}{\inl{V}}{\tysum{T}{U}}}
    \end{mathpar}
  \end{case*}
  The full proof can be found in~\cref{prf:lem-pgv-value-done}.
\end{proof}
\begin{restatablelemma}{lempgvsubstitution}[Substitution]
  \label{lem:pgv-substitution}
  \hfill\\%newline before theorem statement
  If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
\end{restatablelemma}
\begin{proof}
  By induction on the derivation of $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$. Whenever the contexts in the premises and conclusion differ, we use \cref{lem:pgv-value-done} to prove that the priorities are preserved. We demonstrate the case for \LabTirName{T-Lam}:
  \begin{case*}[\LabTirName{T-Lam}]
    By \cref{lem:pgv-value-done}, $\pr(\ty{\Theta})=\pr(\ty{U'})$, hence $\pr(\ty{\Gamma},\ty{\Theta})=\pr(\ty{\Gamma},\ty{U'})$.
    \begin{mathpar}
      \small
      \inferrule*{
        \tseq[\cs{q}]{\ty{\Gamma},\tmty{x}{U'},\tmty{y}{T}}{M}{U}
      }{\tseq[\cs{\pbot}]{\ty{\Gamma},\tmty{x}{U'}}
        {\lambda y.M}
        {\tylolli[\cs{\pr(\ty{\Gamma},\ty{U'})},\cs{q}]{T}{U}}}
      \substarrow{V}{x}
      \inferrule*{
        \tseq[\cs{q}]{\ty{\Gamma},\ty{\Theta},\tmty{y}{T}}{\subst{M}{V}{x}}{U}
      }{\tseq[\cs{\pbot}]{\ty{\Gamma},\ty{\Theta}}
        {\lambda y.\subst{M}{V}{x}}
        {\tylolli[\cs{\pr(\ty{\Gamma},\ty{\Theta})},\cs{q}]{T}{U}}}
    \end{mathpar}
  \end{case*}
  The complete proof can be found in~\cref{prf:lem-pgv-substitution}.
\end{proof}
The proofs for subject congruence and subject reduction are standard inductions. The details can be found in \cref{prf:lem-pgv-subject-reduction-terms,prf:lem-pgv-subject-congruence,prf:thm-pgv-subject-reduction-confs}.
\begin{restatablelemma}{lempgvsubjectreductionterms}[Subject Reduction, $\tred$]
  \label{lem:pgv-subject-reduction-terms}
  \hfill\\%newline before theorem statement
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
  then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
\end{restatablelemma}
\begin{restatablelemma}{lempgvsubjectcongruence}[Subject Congruence, $\equiv$]
  \label{lem:pgv-subject-congruence}
  \hfill\\%newline before theorem statement
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{restatablelemma}
\begin{restatabletheorem}{thmpgvsubjectreductionconfs}[Subject Reduction, $\cred$]
  \label{thm:pgv-subject-reduction-confs}
  \hfill\\%newline before theorem statement
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{restatabletheorem}
